Prediction of Protein-Protein Interactions on the Human and Rice Interactome

### 2020-10-15 nicolas/PPIPL3/ReportV4.tex
Last update on report

Network analyses for predicting unmapped protein-protein interactions
(PPI) suggest that the higher the number of paths of length\textasciitilde 3
(L3) between two proteins, the more likely they are to interact. This
paper extends previous work based on the L3 principle by taking into
account the representation learning of node features of the PPI network.
In particular, we train an \texttt{XGBoost} model using L3 and handcrafted
features, as well as embeddings from \texttt{Node2Vec}. Our main result
shows that while L3 is an important feature for predicting links,
best performance is achieved when combined with edge embeddings. The
proposed approach is evaluated for the human and rice interactomes.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

@@

Based on the outcome of numerous Y2H experiments PPI networks can
be constructed, in which known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
yet unknown interactions. This paper evaluates predictions based on
three measures: Common Neighbors (\textbf{CN}), based on the count
of paths of length 2; the raw count of paths of length 3 (\textbf{A3});
and the degree-normalized count of paths of length 3 (\textbf{L3}).
Furthermore, the Machine Learning (ML) process is assessed for each
metric as the only feature, for the features obtained by the Node2Vec
algorithm and for a combination of the Node2Vec and

@@

The focus of this paper is to evaluate different methods for predicting
PPIs using the existing knowledge of the network of interactions,
which is represented as an undirected graph. The traditional way of
approaching the problem is usually based on social networks analysis,
more specifically on the triadic closure principle (TCP), that states
that the more common shared friends that two people have, the more
likely that they know each other. As shown by previous studies, the
mentioned approach fails because it does not consider the structural
and chemical properties of the proteins \cite{Kovacs2019}.

@@

For the interaction prediction for each network, the algorithms described
below were used. It is important to keep in mind how the protein-protein
interaction (PPI) network $G=(V,E)$ is conceptualized: each node
($v_{i}\in V$) represents a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
represents and interaction among proteins $v_{i}$ and $v_{j}$.
\begin{description}
\item [{Common~Neighbors~(CN)}] This method is based on the Triadic Closure
Principle: ``the more common friends two individuals have, the more
likely that they know each other''. For the implementation of this
method, $A{{}^2}$ matrix is calculated, being $A$ the adjacency
matrix of the network.
\item [{Length-3~Paths~(A3)}] This is the simplest implementation of
the proposed insight of ``if my friends and your friends interact,
then we might interact too''. The calculating is carried on with
$A{{}^3}$, i.e, the third power of the adjacency matrix.
\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
overestimate the importance of some edges due to intermediate hubs
which add many shortcuts in the graph. To address that issue, a degree
normalization for the path $X\rightarrow U\rightarrow V\rightarrow Y$
is applied by considering the degree $k$ of the intermediate nodes
$U$ and $V$, as follows.
\[
p_{XY}=\sum_{U,V}\frac{A_{XU}\cdot A_{UV}\cdot A_{VY}}{\sqrt{k_{U}\cdot k_{V}}}
\]
\\
where $A_{ij}$ represents the value of the adjacency matrix for nodes
$i$ and $j$: 1 if the edge $\{i,j\}$ exists, 0 otherwise.
\end{description}

@@

As mentioned before, 80\% of the final dataset was randomly selected
and used for training, while the remaining 20\% was used for validation.
The whole training-validation procedure was applied 10 times.

@@

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
with each feature}
\par\end{centering}
\noindent \begin{raggedleft}
\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}
\par\end{raggedleft}
\noindent \begin{raggedleft}
\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
\par\end{raggedleft}
\noindent \raggedleft{}\includegraphics[width=0.8\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
\end{figure}

@@

As the first step in the evaluation of CN, A3 and L3 methods for edge
prediction, the full calculation and ranking of all predictions was
carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
and \emph{HI-TESTED}) is shown in Figures \ref{fig:HI1} and \ref{fig:HI2},
in which the x axis represents a rank position $k$ and the y axis
displays the precision for the top $k$ predictions of each method,
when assuming interactome \emph{HI-III} as the validation set.

\begin{figure}[h]
\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14}}

\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-ii-14\lyxdot txt}
\end{figure}

\begin{figure}[h]
\caption{\label{fig:HI2}Methods Comparison for \emph{HI-TESTED}}

\noindent \centering{}\includegraphics[width=1\columnwidth]{hi-tested\lyxdot txt}
\end{figure}

@@

However, it is of our interest to assess if Machine Learning can boost
the overall performance on the prediction, so the proposed procedure
was adapted to the human interactome data: Instead of having to remove
a fraction of edges from the network in order to predict and validate,
now the prediction network and the validation networks are given beforehand.
The rest of the procedure is carried out as mentioned before. Results
for the Interactome \emph{HI-II-14} with Node2Vec and each metric
are presented in Figures \ref{F2-1}, \ref{F3-1} and \ref{F4-1}.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F2-1}\emph{HI-II-14} results for \texttt{Node2Vec} with CN
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F3-1}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F4-1}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
\end{figure}

It can be seen that for \emph{HI-II-14}, L3 performs marginally better
than CN, especially when comparing precision ( vs ) and recall ( vs
), but all of them have good values for area under the ROC curve of
0.99.

@@

>>>>

Network analyses for predicting unmapped protein-protein interactions
(PPI) suggest that the higher the number of paths of length 3 (L3)
between two proteins, the more likely they are to interact. This paper
extends previous work based on the L3 principle by taking into account
the representation learning of node features of the PPI network. In
particular, we train an \texttt{XGBoost} model using L3 and handcrafted
features, as well as embeddings from \texttt{Node2Vec}. Our main result
shows that while L3 is an important feature for predicting links,
best performance is achieved when combined with edge embeddings. The
proposed approach is evaluated for the human and rice interactomes.
\end{abstract}
\begin{comment}
need include a feature that compares the dimension of the embedding
is useful for prediction with L3 to show that they independent measures
- we do not want to be learning L3 through node2vec
\end{comment}

@@

Based on the outcome of numerous Y2H experiments PPI networks can
be constructed, in which known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
yet unknown interactions. The focus of this study is to evaluate different
methods for predicting PPIs using the existing knowledge of the network
of interactions, which is represented as an undirected graph. The
traditional way of approaching the problem is usually based on social
networks analysis, more specifically on the triadic closure principle
(TCP), that states that the more common shared friends that two people
have, the more likely that they know each other. As shown by previous
studies, the mentioned approach fails because it does not consider
the structural and chemical properties of the proteins \cite{Kovacs2019}.

@@

@@

For the interaction prediction for each network, the algorithms described
below were used. It is important to keep in mind how the protein-protein
interaction (PPI) network $G=(V,E)$ is conceptualized: each node
($v_{i}\in V$) represents a protein and each undirected edge ($e_{b}=\{v_{i},v_{j}\},\,e_{b}\in E$)
represents and interaction among proteins $v_{i}$ and $v_{j}$.
\begin{description}
\item [{Common~Neighbors~(CN)}] This method is based on the Triadic Closure
Principle: ``the more common friends two individuals have, the more
likely that they know each other''. For the implementation of this
method, $A{{}^2}$ matrix is calculated, being $A$ the adjacency
matrix of the network.
\item [{Length-3~Path~Counts~(A3)}] This is the simplest implementation
of the proposed insight of ``if my friends and your friends interact,
then we might interact too''. The calculating is carried on with
$A{{}^3}$, i.e, the third power of the adjacency matrix.
\item [{Degree-normalized~L-3~Score~(L3)}] The previous approach might
overestimate the importance of some edges due to intermediate hubs
which add many shortcuts in the graph. To address that issue, a degree
normalization for the path $X\rightarrow U\rightarrow V\rightarrow Y$
is applied by considering the degree $k$ of the intermediate nodes
$U$ and $V$, as follows.
\[
p_{XY}=\sum_{U,V}\frac{A_{XU}\cdot A_{UV}\cdot A_{VY}}{\sqrt{k_{U}\cdot k_{V}}}
\]
\\
where $A_{ij}$ represents the value of the adjacency matrix for nodes
$i$ and $j$: 1 if the edge $\{i,j\}$ exists, 0 otherwise.
\end{description}

@@

As mentioned before, 80\% of the final dataset was randomly selected
and used for training, while the remaining information was used for
validation. The whole training-validation procedure was applied 10
times.

@@

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F8-importance}Importance gain plots for \texttt{Node2Vec}
with each feature}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.7\columnwidth]{ML_Metric/Imp\lyxdot A2\lyxdot All}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.7\columnwidth]{ML_Metric/Imp\lyxdot A3\lyxdot All}
\par\end{centering}
\centering{}\includegraphics[width=0.7\columnwidth]{ML_Metric/Imp\lyxdot L3\lyxdot All}
\end{figure}

@@

As the first step in the evaluation of CN, A3 and L3 methods for edge
prediction, the full calculation and ranking of all predictions was
carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
the x axis represents a rank position $k$ and the y axis displays
the precision for the top $k$ predictions of each method, when assuming
interactome \emph{HI-III} as the validation set.

\begin{figure}[h]
\caption{\label{fig:HI1}Methods Comparison for \emph{HI-II-14} and \emph{HI-TESTED}}

\noindent \centering{}\includegraphics[width=0.48\columnwidth]{hi-ii-14\lyxdot txt}\includegraphics[width=0.48\columnwidth]{hi-tested\lyxdot txt}
\end{figure}

@@

However, it is of our interest to assess if Machine Learning can boost
the overall performance on the prediction, so the proposed procedure
was adapted to the human interactome data: Instead of having to remove
a fraction of edges from the network in order to predict and validate,
now the prediction network and the validation networks are given beforehand.
The rest of the procedure is carried out as mentioned before. Results
for the Interactome \emph{HI-II-14} with Node2Vec and each metric
are presented in Figures \ref{HI1}, \ref{HI2} and \ref{HI3}.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI1}\emph{HI-II-14} results for \texttt{Node2Vec} with CN
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI2}\emph{HI-II-14} results for \texttt{Node2Vec} with A3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI3}\emph{HI-II-14} results for \texttt{Node2Vec} with L3
feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3}
\end{figure}

It can be seen that for \emph{HI-II-14}, CN performs marginally better
than L3 when comparing precision ($0.9978$ vs $0.9977$) but worse
in terms of recall ($0.9346$ vs $0.9598$), although all metrics
have a good values for area under the ROC curve of 0.99.

On the other hand, the interactome \emph{HI-TESTED} was assessed with
the same methodology as \emph{HI-II-14} and results are shown in Figures
\ref{HI1-1}, \ref{HI2-1} and \ref{HI3-1}. It can be seen that in
terms of false positives, CN performs better than A3 and L3, but in
terms of false negatives, A3 actually performs better that CN and
L3. In terms of AUC, metrics results are indistinguishable.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI1-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
CN feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A2\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A2\lyxdot 2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI2-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
A3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot A3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot A3\lyxdot 2}
\end{figure}

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{HI3-1}\emph{HI-TESTED} results for \texttt{Node2Vec} with
L3 feature}
\par\end{centering}
\noindent \raggedleft{}\includegraphics[width=0.48\columnwidth]{Human/ROC\lyxdot L3\lyxdot 2}\includegraphics[width=0.48\columnwidth]{Human/CM\lyxdot L3\lyxdot 2}
\end{figure}

An interesting point from this human datasets can be observed when
looking at the importance plots, that show a very biased influence
of the handcrafted features. Figure X shows this behavior for the
\emph{HI-II-14} dataset, which is very similar to \emph{HI-TESTED}.

\begin{figure}[H]
\noindent \begin{centering}
\caption{\label{F8-importance-1}Importance gain plots for \emph{HI-II-14}}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A2\lyxdot 1}
\par\end{centering}
\begin{centering}
\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot A3\lyxdot 1}
\par\end{centering}
\centering{}\includegraphics[width=0.7\columnwidth]{Human/Imp\lyxdot L3\lyxdot 1}
\end{figure}

@@

It was found that applying \texttt{XGBoost} to the rice PPI prediction
obtained very good results when only using the node embedding technique
(AUC=0.93) and when only using the handcrafted features based on protein
neighborhood (AUC \textasciitilde 0.88). Even better results were
achieved when the handcrafted features and the vector representation
information were combined (AUC \textasciitilde 0.98). It is important
to consider that the importance gain of the handcrafted feature in
each model was high, although importance values for the A3 method
stand out less from the node embedding information than in CN and
L3.

On the human interactome, the results show that on both interactomes
(\emph{HI-II-14} and \emph{HI-TESTED}) all metrics yield very good
results for the link prediction problem, and therefore this study
extends previous work by other authors that only consider the predictive
power of the metrics alone.

It is important to note the difference in feature importance from
rice and human interactomes. In the former, features have a more balanced
importance distribution than in the latter dataset, whose model shows
a very strong dependence on the handcrafted feature.

With this study, a framework for link prediction on protein-protein
interactions is presented and validated against the available information
on rice and human interactomes, but can be easily extended to other
organisms or even to other graphs not necessarily related to biology.
